Summary of cell type ontologies in reference files

### 2024-12-17 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
keep myeloid and epithelial

```{r}
# turn cl_ont into data frame with one row per term 
cl_df <- data.frame(
  cl_ontology = cl_ont$id,
  cl_annotation = cl_ont$name
) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # list all ancestors and descendants calculate total
    ancestors = list(ontologyIndex::get_ancestors(cl_ont, cl_ontology)),
    total_ancestors = length(ancestors),
    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology)),
    total_descendants = length(descendants)
  )
```

@@

```{r}
# first set up the graph from cl ont
parent_terms <- cl$parents
cl_graph <- igraph::make_graph(rbind(unlist(parent_terms), rep(names(parent_terms), lengths(parent_terms))))
```

@@

I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte.

@@

Same with `progenitor cell`, I do think it could be helpful to know that something may be a progenitor cell, but when you have a cell with the label for HSC and the label for cells like monocytes or osteoblasts, then maybe we are talking about a tumor cell instead.

@@

The only term in this list that I would be concerned about losing is "neuron". 
Let's look at those combinations.

@@

```{r}
# blood cell
print_df |> 
  dplyr::filter(cl_annotation == "neuron")
```

@@

@@

- Pairs should not have more than 1 LCA, with the exception of the matches that have the label hematopoietic precursor cell. 
- The LCA should have equal to or less than 170 total descendants. 
- We whould include the term for `neuron` even though it has 500 descendants. 
- Terms that are too broad (like `supporting cell`, `blood cell`, `bone cell`, `lining cell`) should be removed.

>>>>

```{r}
# turn cl_ont into data frame with one row per term 
cl_df <- data.frame(
  cl_ontology = cl_ont$id,
  cl_annotation = cl_ont$name
) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # list all ancestors and descendants calculate total
    ancestors = list(ontologyIndex::get_ancestors(cl_ont, cl_ontology)),
    total_ancestors = length(ancestors),
    descendants = list(ontologyIndex::get_descendants(cl_ont, cl_ontology, exclude_roots = TRUE)),
    total_descendants = length(descendants)
  )
```

@@

```{r}
# first set up the graph from cl ont
parent_terms <- cl_ont$parents
cl_graph <- igraph::make_graph(rbind(unlist(parent_terms), rep(names(parent_terms), lengths(parent_terms))))
```

@@

I'm torn on this one, because I do think it's helpful to know if something is of the myeloid lineage, but if we aren't keeping lymphocyte then I would argue we shouldn't keep myeloid leukocyte. 
Noting that after discussion we have decided to keep this one since T and B cells are much easier to differentiate based on gene expression alone than cells that are party of the myeloid lineage.

@@

Same with `progenitor cell`, I do think it could be helpful to know that something may be a progenitor cell, but when you have a cell with the label for HSC and the label for cells like monocytes or osteoblasts, then maybe we are talking about a tumor cell instead. 
After discussion, we are going to remove progenitor cells.

@@

The only terms in this list that I would be concerned about losing are "neuron" and epithelial cells. 
Let's look at those combinations.

@@

```{r}
# neuron
print_df |> 
  dplyr::filter(cl_annotation == "neuron")
```

@@

#### Epithelial cell

```{r}
# epithelial cell
print_df |> 
  dplyr::filter(cl_annotation == "epithelial cell")
```

The PanglaoDB cell types seem to be more specific than the ones present in Blueprint Encode, similar to the observation with neurons.
We should keep epithelial cell.

@@

- Pairs should not have more than 1 LCA, with the exception of the matches that have the label hematopoietic precursor cell. 
- The LCA should have equal to or less than 170 total descendants. 
- We whould include the term for `neuron` and `epithelial cell` even though they do not pass the threshold for number of descendants. 
- Terms that are too broad should be removed. 
This includes: `lining cell`, `blood cell`, `progenitor cell`, `bone cell`, and `supporting cell`
